---
title: "Assignment 2: Git and Stream Temperature - July Analysis"
author: "Grace Bryan"
date: "May 07, 2026"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

### **Data Loading**

```{r}
library(tidyverse)
library(lubridate)

# Load saved data

temp_data <- read.csv("data/temp_data.csv")

# Select and rename relevant columns 
temp_data <- temp_data %>% 
  select(
    agency_cd = monitoring_location_id, 
    site_no = monitoring_location_id, 
    Date = time, 
    temp_C = value ) 

# Clean site names
temp_data$site_no <- gsub("USGS-", "", temp_data$site_no) 

# Ensure proper format 
temp_data$Date <- as.Date(temp_data$Date) 
temp_data$month <- month(temp_data$Date, label = TRUE)
temp_data$year <- year(temp_data$Date)

```

### **Filter for July**

```{r}
aug_data <- temp_data %>%
  filter(month == "Aug")
```

### **Model: Temperature Trend Over Time**

We use a generalized liner model (GLM) to estimate whether stream temperature is chaning over time and whether trends differ across sites.

```{r}
aug_model <- glm(temp_C ~ year * site_no,
                  data = aug_data)

summary(aug_model)
```

\###*Visualization of Trends*

```{r}
ggplot(aug_data, aes(x = year, y = temp_C,
                      color = site_no)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "August Temperature Trends by Site",
    x = "Year",
    y = "Temperature (˚C)" ) +
  theme_minimal()
```

### **Interpretation of Results**

August stream temperatures show a clear increasing trend over time. The coefficient for year is positive and highly significant (p \< 0.001), indicating that temperatures are rising at a rate of approximately 0.13˚C per year at the baseline site (site 14166000).

There are also differences in warming rates between sites. Site 14211010 shows a significantly greater increase in temperature over time compared to the baseline site, with an additional warming rate of about 0.046˚C per year. This suggests that this site is warming faster than the baseline.

In contrast, site 14211720 does not show a statistically significant difference in warming rate compared to the baseline site. Therefore, there is no strong evidence that this site is warming at a different rate.

Overall, the results indicate that August stream temperatures are increasing over time, with some variation in warming rates across sites.
