---
title: "Assignment 2: Git and Stream Temperature - Data Download"
author: "Grace Bryan"
date: "May 07, 2026"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

### **Data Source**

Daily stream temperature data were downloaded from the USGS NWIS database for three sites on the Willamette River from 2010-2023.

### **Data Download or Load**

```{r}

library(dataRetrieval)
library(tidyverse)
library(lubridate)

# Site IDs (USGS format required for API)
siteNumbers <- c("USGS-14211720", "USGS-14211010", "USGS-14166000")

# Toggle: TRUE = download fresh data, FALSE = load saved data

download_data <- TRUE

if(download_data){

  temp_data <- read_waterdata_daily(
    monitoring_location_id = siteNumbers,
    parameter_code = "00010", # Water temp
    statistic_id = "00003",   # Daily mean
    time = c("2010-01-01", "2023-12-31"),
    skipGeometry = TRUE
  )

# Save data locally
write.csv(temp_data, "data/temp_data.csv", row.names = FALSE)

} else {

  temp_data <- read.csv("data/temp_data.csv")
}
```

### **Data Processing**

The raw dataset was cleaned and reshaped for analysis. Relevant columns were selected and renamed for clarity, and time variables (month and year) were extracted to support analysis.

```{r}

# Select and rename relevant columns 
temp_data <- temp_data %>% 
  select(
    agency_cd = monitoring_location_id, 
    site_no = monitoring_location_id, 
    Date = time, 
    temp_C = value ) 

# Clean site names
temp_data$site_no <- gsub("USGS-", "", temp_data$site_no) 

# Ensure Date format 
temp_data$Date <- as.Date(temp_data$Date) 

# Extract time variables 
temp_data$month <- month(temp_data$Date, label = TRUE)
temp_data$year <- year(temp_data$Date)

```

### **Visualization**

The plot below shows temperature variation over time for each monitoring site.

```{r}
# Plot
ggplot(temp_data, aes(x = Date, y = temp_C, color = site_no)) + 
  geom_line() + 
  labs(  
    title = "Stream Temperature at Willamette River Sites", 
    x = "Date", 
    y = "Temperature (°C)", 
    color = "Site Number" ) + 
  theme_minimal()
```

### *Summary Statistics*

```{r}
temp_data %>%
  group_by(site_no) %>%
  summarise(
    mean_temp = mean(temp_C, na.rm = TRUE),
    max_temp = max(temp_C, na.rm = TRUE),
    min_temp = min(temp_C, na.rm = TRUE)
  )
```
